---
title: "Assignment 3: Data Exploration"
author: "Ariel Lam"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai.

The completed exercise is due on Sept 30th.

## Set up your R session

1.  Check your working directory, load necessary packages (tidyverse), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r,  message = FALSE}
getwd()
library(tidyverse)

Neonics <- read.csv("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", stringsAsFactors = TRUE)
Litter <- read.csv("./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv", stringsAsFactors = TRUE)

```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Considering Neonicotinoids are a class of insecticide it is necessary to consider its impacts on insects to assess how toxic these substances are. This is after all because insecticides are used to kill off insects and in order for them to be effective, they must reach a certain toxicity threshold. The study of these thresholds are imperative as insecticides could prove harmful to beneficial insects such as bees, or fail to serve their purpose in killing their target organisms. 

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Might be interested in assessing the litter and woody debris gauge the different patterns of tree sheddings or their diversities to learn more about certain seasonal/ plant behavior.

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. Litter is randomly sampled from tower plots within the 90% flux footprint of the primary and secondary airsheds. 
 2. Buffer around plots to avoid interference.
 3. Trap placement is either  targeted or randomized depending on the target vegetation. 

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r}
dim(Neonics)
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?

```{r}
summary(Neonics$Effect)
```

> Answer: These might be of interest as it reflects the behavior of the insecticide on insects and the potential effectiveness of Neonicotinoids on agricultural uses. 

7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.

```{r}
summary(Neonics$Species.Common.Name) 
```

> Answer: Honey Bee, Parasitic Wasp, Buff Tailed Bumblebee, Carniolan Honey Bee, Bumble Bee, and the Italian Honeybee are the most common. Bees have been reported to be the most vulnerable to insecticides which is detrimental to ecosystems as they are key pollinators. Their disappearance wreaks havoc on the earth. 

8.  Concentrations are always a numeric value. What is the class of Conc.1..Author. in the dataset, and why is it not numeric?

```{r}
class(Neonics$Conc.1..Author.)
```

> Answer: The class is a factor. This is not numeric due to the fact that all the concentrations are reported in different units. 

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r}
ggplot(Neonics, aes(Publication.Year))+ geom_freqpoly()
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r}
ggplot(Neonics, aes(Publication.Year, colour= Test.Location))+ geom_freqpoly()

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer:The most common test location looks to be in the lab and the natural field. It looks as though there is an increase in lab location over time and a gradual increase for the field, but a descrease in the recent years. 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information.

```{r}
ggplot(Neonics, aes(y=Endpoint))+ geom_bar()

```

> Answer: Most common is NOEL and LOEL. NOEL stands for "No-observable-effect-level: highest dose (concentration) producing effects not significantly different from responses of controls according to author's reported statistical tests". LOEL stands for "lowest-observable-effect-level: lowest (concentration) producing effects that are significantly different". 


## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r}
class(Litter$collectDate)
Litter$collectDate <- as.Date(Litter$collectDate, format="%Y-%m-%d")


```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
unique(Litter$plotID)
#summary(Litter$plotID)
```

> Answer: 12 plot IDs. The unique function identifies the distinct categories within the column whilst the summary generates the count within each distinct categories

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r}
ggplot(Litter, aes(y=functionalGroup)) +geom_bar()

```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r}
ggplot(Litter, aes(y=dryMass, x=functionalGroup)) +geom_boxplot()

ggplot(Litter, aes(y= dryMass, x=functionalGroup)) +geom_violin()


```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer:The boxplots display outliers and the quartiles better than the violin plot. It is a lot easier to distinguish between the dryMass values compared to functional group in the boxplots and compare them than it is to the violine plot which does not show much. 

What type(s) of litter tend to have the highest biomass at these sites?

> Answer:The Needles tend to have the highest biomass at these sites

