---
title: "Assignment 2: Coding Basics"
author: "Ariel Lam"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r}
#1. 
Sequence_of_four <- seq(1,100,4) #assign sequence to a variable
#2. 
mean(Sequence_of_four) #calculate the mean of the variable/ sequence
median(Sequence_of_four) #calculate the median of the variable/ sequence
#3. 
mean(Sequence_of_four) > median(Sequence_of_four) #ask if the mean of the sequence is greater than the median. boolean result.
```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r}
name <- c("Bob", "Rob", "Cob", "Gob") #character vector
test_score <- c(5,20,8,99) # integer vector
pass <- c(FALSE, FALSE, FALSE, TRUE) #logical vector
class_scores <- data.frame(name, test_score, pass)
names(class_scores)[names(class_scores) == 'name'] <- 'Student Name'
names(class_scores)[names(class_scores) == 'test_score'] <- 'Test Score'
names(class_scores)[names(class_scores) == 'pass'] <- 'Pass?'


```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: Data frame is different because it can store different classes of data. In this case, the data frame stores three different classes of data (logical, character, and integer). A matrix will only stores one class. 

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r}

passing_grade <- function(x) {
  ifelse(x<50, "True", "False")}

passing_grade(test_score)

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: ifelse worked as it can pass vectors. The separate 'if' and 'else' functions can only pass separate integers. 
