---
title: "Assignment 5: Data Visualization"
author: "Ariel Lam"
output: pdf_document
geometry: margin=2.54cm
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=45), tidy=TRUE)
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

The completed exercise is due on Friday, Oct 14th @ 5:00pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse, 
lubridate, & cowplot packages. Upload the NTL-LTER processed data files for  
nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] version) and 
the processed data file for the Niwot Ridge litter dataset (use the
[`NEON_NIWO_Litter_mass_trap_Processed.csv`] version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}

#1 
getwd()
setwd("/Users/ariellam/Desktop/EDA-Fall2022")
library(tidyverse)
library(lubridate)
library(cowplot)

PeterPaul <- read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors = TRUE)
NiwotRidge <- read.csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", stringsAsFactors = TRUE)
#2 

PeterPaul$sampledate<- as.Date(PeterPaul$sampledate, format = "%Y-%m-%d")
NiwotRidge$collectDate<- as.Date(NiwotRidge$collectDate, format = "%Y-%m-%d")


```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#3
assignment_theme <- theme_light(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right")

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best 
practices for data visualization. Ensure your theme, color palettes, axes, 
and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with 
separate aesthetics for Peter and Paul lakes. Add a line of best fit and 
color it black. Adjust your axes to hide extreme values (hint: change the 
limits using `xlim()` and/or `ylim()`).

```{r}
#4
ggplot(PeterPaul, aes(x=po4, y =tp_ug))+ assignment_theme +geom_point(aes(color=lakename))+xlim(c(0,55))+ geom_smooth(method=lm, color="black")+ylab("Total Phosphorus (ug)")+xlab("Phosphate")+ggtitle("Total Phosphorus and Phosphate in Peter and Paul Lakes")

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and 
(c) TN, with month as the x axis and lake as a color aesthetic. Then, create 
a cowplot that combines the three graphs. Make sure that only one legend is 
present and that graph axes are aligned. 

Tip: R has a build in variable called `month.abb` that returns a list of months;
see <https://r-lang.com/month-abb-in-r-with-example>

```{r}
#5
temperature_boxplot <-
  ggplot(PeterPaul, aes(x = month, y = temperature_C)) +
  geom_boxplot(aes(color = lakename))+ assignment_theme+ylab("Temperature (C)")+xlab("Month")+ggtitle("Temperature Distribution")
print(temperature_boxplot)

TP_boxplot <-
  ggplot(PeterPaul, aes(x = month, y = tp_ug)) +
  geom_boxplot(aes(color = lakename))+ assignment_theme+ylab("TP (ug)")+xlab("Month")+ggtitle("Total Phosphorus (ug) Distribution")
print(TP_boxplot)

TN_boxplot <-
  ggplot(PeterPaul, aes(x = month, y = tn_ug)) +
  geom_boxplot(aes(color = lakename))+ assignment_theme+ylab("TN (ug)")+xlab("Month")+ggtitle("Total Nitrogen (ug) Distribution")
print(TN_boxplot)

plot_grid(temperature_boxplot, TP_boxplot, TN_boxplot, nrow = 3, align = 'h', rel_heights = c(1, 1))
```

Question: What do you observe about the variables of interest over seasons 
and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the 
"Needles" functional group. Plot the dry mass of needle litter by date and 
separate by NLCD class with a color aesthetic. (no need to adjust the name 
of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into 
three facets rather than separated by color. 

```{r}
#6

#7


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: