---
title: "Assignment 09: Data Scraping"
author: "Ariel Lam"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A09_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Check your working directory
* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Set your ggplot theme

```{r, message = FALSE}
#1
getwd()
library(tidyverse)
library(rvest)
library(lubridate)
library(viridis)
library(dataRetrieval)
library(tidycensus)
library(formatR)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=45), tidy=TRUE)


mytheme <- theme_classic() +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2021 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2021>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2
durham_municipality <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2021')
durham_municipality

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PSWID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Daily Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings), with the first value being "27.6400".

```{r scrape.the.data}
#3
water.system.name <- durham_municipality %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
  
pswid <- durham_municipality %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
  
ownership <- durham_municipality %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
  
max.withdrawals.mgd <- durham_municipality %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc...

5. Create a line plot of the maximum daily withdrawals across the months for 2021

```{r create.a.dataframe.from.scraped.data}
#4
durham_withdrawals <- data.frame("Name" = rep(water.system.name),
                             "PSWID" = rep(pswid),
                             "Ownership" = rep(ownership), 
                             "Month" = month(c(1, 5,9, 2, 6,10, 3, 7, 11,4, 8, 12)),
                             "Year"= 2021,
                             "Max_Withdrawals_mgd" = as.numeric(max.withdrawals.mgd))
durham_withdrawals <-durham_withdrawals %>% mutate ("Date"= my(paste0(Month,"-",Year)))

#5


ggplot(durham_withdrawals,aes(x = Date, y=Max_Withdrawals_mgd, group=1)) + 
  geom_line() + geom_smooth(method="loess",se=FALSE) +
  labs(title = "2021 Water usage data for Durham",
       y="Withdrawal (mgd)",
       x="Date")

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6.
scrape.it <- function(the_year, the_site){
  
  #Retrieve the website contents 
  the_website <- read_html(paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=', the_site, '&year=', the_year))
  
  water_system_tag <- 'div+ table tr:nth-child(1) td:nth-child(2)'
  site_tag <- 'td tr:nth-child(1) td:nth-child(5)'
  ownership_tag <- 'div+ table tr:nth-child(2) td:nth-child(4)'
  the_data_tag <- 'th~ td+ td'
  
  #Scrape the data items
  the_water_system <- the_website %>% html_nodes(water_system_tag) %>% html_text()
  site_name <- the_website %>%   html_nodes(site_tag) %>%  html_text()
  ownership_type <- the_website %>% html_nodes(ownership_tag) %>% html_text()
  max_withdrawals <- the_website %>% html_nodes(the_data_tag) %>% html_text()
  
  #Convert to a dataframe
  df_withdrawals <- data.frame("Name" = rep(the_water_system),
                             "PSWID" = rep(site_name),
                             "Ownership" = rep(ownership_type), 
                             "Month" = month(c(1, 5,9, 2, 6,10, 3, 7, 11,4, 8, 12)),
                             "Year"= the_year,
                             "Max_Withdrawals_mgd" = as.numeric(max_withdrawals))
 df_withdrawals <- df_withdrawals %>% mutate ("Date"= my(paste0(Month,"-",Year)))
   
  
  #Return the dataframe
  return(df_withdrawals)
}


```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7
durham_2015 <- scrape.it(2015,'03-32-010')
view(durham_2015)

ggplot(durham_2015,aes(x = Date, y=Max_Withdrawals_mgd, group=1)) + 
  geom_line() + geom_smooth(method="loess",se=FALSE) +
  labs(title = "2015 Water usage data for Durham",
       y="Withdrawal (mgd)",
       x="Date")

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8
ashville_2015 <- scrape.it(2015,'01-11-010')
view(ashville_2015)

combine_2015 <- rbind(ashville_2015,durham_2015)


ggplot(combine_2015, aes(x = Date, y=Max_Withdrawals_mgd, group=Name)) + geom_line(aes(color=Name))+labs(title = "2015 Water usage data for Durham and Ashville",
       y="Withdrawal (mgd)",
       x="Date")

```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2019.Add a smoothed line to the plot.

>TIP: See Section 3.2 in the "09_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data}
#9

the_years <- c(2010,2011,2012,2013,2014,2015,2016,2017,2018,2019)

asheville_2010_2019 <- map2(the_years, '01-11-010',scrape.it) %>% bind_rows()

#asheville_2010_2019_dataframe <- bind_rows(asheville_2010_2019)

ggplot(asheville_2010_2019, aes(x = Date, y=Max_Withdrawals_mgd)) + geom_line()+  geom_smooth(method="loess",se=FALSE) +labs(title = "2010-2019 Water usage data for Ashville",
       y="Withdrawal (mgd)",
       x="Date")
```

>Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time? The water usage for Ashville seems to be gradually increasing over time. 


